A tutorial on Bayesian and Frequentist Event History Analyses for psychological time-to-event data

### 2024-09-21 manuscript/manuscript.tex
update manuscript

\abstract{%
Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. The orthodox method for analysing such data -- comparing means with analysis-of-variance -- is actually hiding a lot of information about psychological effects, such as their onset time and duration, and how they evolve with increasing waiting time. Such information can change key conclusions about psychological processes and can be revealed by using distributional measures that portray the detailed shape of time-to-event distributions.\\
Here we provide a set of tutorials on how to implement one particular distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. We illustrate how one can calculate the descriptive statistics, and how one can implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. The R code is publicly available on Github and OSF, and can easily be adapted for other data sets. We discuss possible link functions and prior distributions, how to manage inter-individual differences, implications for experimental design, and the advantages of a hazard analysis over other distributional methods available in the literature. Our ultimate goal is to convince readers to start using hazard analysis more often when dealing with time-to-event data.
}

@@

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in means conceal when an experimental effect starts, how long it lasts, how it evolves over increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).
As a simple illustration, Figure 1 shows three examples of how an observed difference in mean response times (RTs) between two experimental conditions conceals differences in the shapes of the underlying RT distributions.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

These results show that this participant is initially responding to the prime even though (s)he was instructed to only respond to the target, that response competition emerges in the incongruent prime condition around 300 ms, and that only later response are fully controlled by the target stimulus. Qualitatively similar results were obtained for the other five participants. Also, in their second Experiment, Panis and Schmidt (2016) showed that the negative compatibility effect in the mask-present conditions is time-locked to mask onset. This example shows that a simple difference between two means fails to reveal the dynamic behavior people display in many experimental paradigms (Panis, 2020; Panis, Moran, Wolkersdorfer, \& Schmidt, 2020; Panis \& Schmidt, 2022; Panis, Torfs, Gillebert, Wagemans, \& Humphreys, 2017; Panis \& Wagemans, 2009; Schmidt, Panis, Wolkersdorfer, \& Vorberg, 2022). In other words, statistically controlling for the passage of time during data analysis is equally important as experimental control during the design of an experiment, to better understand human behavior in experimental paradigms. As we will show in Tutorials 2 and 3, statistical models for h(t) can be implemented as generalized linear mixed regression models predicting event occurrence (1/0) in each bin of a selected time range.

@@

In this tutorial we will fit four Bayesian multilevel models (i.e., generalized linear mixed models) to the data. We select the analysis range (200,600{]} and the cloglog link. The data is prepared as follows.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# load person{-}trial{-}bin oriented data set}
\CommentTok{\#ptb\_data \textless{}{-} read\_csv("Tutorial\_1\_descriptive\_stats/data/inputfile\_hazard\_modeling.csv")}
\CommentTok{\# select analysis time range: (200,600] with 10 bins (time bin ranks t=6 to t=15) of 40 ms}
\CommentTok{\#ptb\_data \textless{}{-} ptb\_data \%\textgreater{}\% filter(period \textgreater{} 5)}
\CommentTok{\# create factor condition, with "blank" as the reference level}
\CommentTok{\#ptb\_data \textless{}{-} ptb\_data \%\textgreater{}\% mutate(condition = factor(condition, labels = c("blank", "congruent","incongruent")))}
\CommentTok{\# center discrete TIME (period) on bin 9, and trial on trial 1000}
\CommentTok{\#ptb\_data \textless{}{-} ptb\_data \%\textgreater{}\% mutate(period\_9 = period {-} 9,}
  \CommentTok{\#                              trial\_c = (trial {-} 1000)/1000)}
\CommentTok{\# remove unnecessary columns before fitting a model}
\CommentTok{\#ptb\_data \textless{}{-} ptb\_data \%\textgreater{}\% select({-}c(bl,tr,trial,period)) \# 12840 obs. of 5 variables}
\end{Highlighting}
\end{Shaded}

@@

@@

For the first model, we use a general specification of TIME (i.e., one intercept per time bin) for the baseline condition (blank prime), and assume that the effects of prime-target congruency and trial number are proportional and additive, and that the effect of trial number is linear.

@@

, and use a polynomial to specify the effect of TIME in the ``blank'' prime condition.

@@

\subsection{Advantages of hazard functions}\label{advantages-of-hazard-functions}

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by Holden, Van Orden, and Turvey (2009), ``probability density functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions'' (p.~331; see also Figure 1 in \ldots.) when one is interested in studying the detailed shape of a RT distribution.

@@

One important limitation is that you need a relatively large number of trials per condition to estimate the hazard function with high temporal resolution. There is a tradeoff between the number of trials per condition and the temporal resolution (i.e., bin width) of the hazard function. We therefore recommend to design as many trials as possible per experimental condition given the available resources.
Another limitation of multilevel hazard modeling is that it might take some time to fit the model.

>>>>

\abstract{%
Time-to-event data such as response times, saccade latencies, and fixation durations are ubiquitous in experimental psychology. The orthodox method for analysing such data -- comparing means with analysis-of-variance -- is actually hiding a lot of information about psychological effects, such as their onset time and duration, and how they evolve with increasing waiting time. Such information can change key conclusions about psychological processes and can be revealed by using distributional measures that portray the detailed shape of time-to-event distributions.\\
Here we provide a set of tutorials on how to implement one particular distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis. We illustrate how one can calculate the descriptive statistics, and how one can implement Bayesian and frequentist regression models, using the R packages tidyverse, brms, and lme4. The R code is publicly available on Github and OSF, and can easily be adapted for other data sets. We discuss possible link functions and prior distributions, how to manage inter-individual differences, implications for experimental design, the advantages of a hazard analysis over other distributional methods available in the literature, limitations, and extensions. Our ultimate goal is to convince readers to start using hazard analysis more often when dealing with time-to-event data.
}

@@

In experimental psychology, it is still standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves over increasing waiting time, and whether its onset is time-locked to other events. Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).
As a simple illustration, Figure 1 shows three examples of how an observed difference in mean response times (RTs) between two experimental conditions conceals differences in the shapes of the underlying RT distributions.

@@

We used R (Version 4.4.0; R Core Team, 2024)\footnote{We, furthermore, used the R-packages \emph{bayesplot} (Version 1.11.1; Gabry, Simpson, Vehtari, Betancourt, \& Gelman, 2019), \emph{brms} (Version 2.21.0; Bürkner, 2017, 2018, 2021), \emph{citr} (Version 0.3.2; Aust, 2019), \emph{cmdstanr} (Version 0.8.1.9000; Gabry, Češnovar, Johnson, \& Bronder, 2024), \emph{dplyr} (Version 1.1.4; Wickham, François, Henry, Müller, \& Vaughan, 2023), \emph{forcats} (Version 1.0.0; Wickham, 2023a), \emph{ggplot2} (Version 3.5.1; Wickham, 2016), \emph{lme4} (Version 1.1.35.5; Bates, Mächler, Bolker, \& Walker, 2015), \emph{lubridate} (Version 1.9.3; Grolemund \& Wickham, 2011), \emph{Matrix} (Version 1.7.0; Bates, Maechler, \& Jagan, 2024), \emph{nlme} (Version 3.1.166; Pinheiro \& Bates, 2000), \emph{papaja} (Version 0.1.2.9000; Aust \& Barth, 2023), \emph{patchwork} (Version 1.2.0; Pedersen, 2024), \emph{purrr} (Version 1.0.2; Wickham \& Henry, 2023), \emph{RColorBrewer} (Version 1.1.3; Neuwirth, 2022), \emph{Rcpp} (Eddelbuettel \& Balamuta, 2018; Version 1.0.12; Eddelbuettel \& François, 2011), \emph{readr} (Version 2.1.5; Wickham, Hester, \& Bryan, 2024), \emph{RJ-2021-048} (Bengtsson, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021, 2021), \emph{standist} (Version 0.0.0.9000; Girard, 2024), \emph{stringr} (Version 1.5.1; Wickham, 2023b), \emph{tibble} (Version 3.2.1; Müller \& Wickham, 2023), \emph{tidybayes} (Version 3.0.6; Kay, 2023), \emph{tidyr} (Version 1.3.1; Wickham, Vaughan, \& Girlich, 2024), \emph{tidyverse} (Version 2.0.0; Wickham et al., 2019), and \emph{tinylabels} (Version 0.2.4; Barth, 2023).} for all reported analyses. Web links are printed in bold. The content of the tutorials is mainly based on Allison (2010), Singer and Willett (2003), McElreath (2018), Kurz (2023a), and Kurz (2023b).

@@

These results show that this participant is initially responding to the prime even though (s)he was instructed to only respond to the target, that response competition emerges in the incongruent prime condition around 300 ms, and that only later response are fully controlled by the target stimulus. Qualitatively similar results were obtained for the other five participants. These results go against the (often implicit) assumption that all observed responses are primed responses to the target stimulus.

Also, in their second Experiment, Panis and Schmidt (2016) showed that the negative compatibility effect in the mask-present conditions (see Tutorial 4) is time-locked to mask onset. This example shows that a simple difference between two means fails to reveal the dynamic behavior people display in many experimental paradigms (Panis, 2020; Panis, Moran, Wolkersdorfer, \& Schmidt, 2020; Panis \& Schmidt, 2022; Panis, Torfs, Gillebert, Wagemans, \& Humphreys, 2017; Panis \& Wagemans, 2009; Schmidt, Panis, Wolkersdorfer, \& Vorberg, 2022). In other words, statistically controlling for the passage of time during data analysis is equally important as experimental control during the design of an experiment, to better understand human behavior in experimental paradigms. As we will show in Tutorials 2 and 3, statistical models for h(t) can be implemented as generalized linear mixed regression models predicting event occurrence (1/0) in each bin of a selected time range.

@@

In this tutorial we will fit four Bayesian multilevel models (i.e., generalized linear mixed models) to the person-trial-bin oriented data set that we created in Tutorial 1. We select the analysis range (200,600{]} and the cloglog link. The data is prepared as follows.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# load person{-}trial{-}bin oriented data set}
\NormalTok{ptb\_data }\OtherTok{\textless{}{-}} \FunctionTok{read\_csv}\NormalTok{(}\StringTok{"../Tutorial\_1\_descriptive\_stats/data/inputfile\_hazard\_modeling.csv"}\NormalTok{)}

\CommentTok{\# select analysis time range: (200,600] with 10 bins (time bin ranks 6 to 15)}
\NormalTok{ptb\_data }\OtherTok{\textless{}{-}}\NormalTok{ ptb\_data }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(period }\SpecialCharTok{\textgreater{}} \DecValTok{5}\NormalTok{)}

\CommentTok{\# create factor condition, with "blank" as the reference level}
\NormalTok{ptb\_data }\OtherTok{\textless{}{-}}\NormalTok{ ptb\_data }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{mutate}\NormalTok{(}\AttributeTok{condition =} \FunctionTok{factor}\NormalTok{(condition, }\AttributeTok{labels =} \FunctionTok{c}\NormalTok{(}\StringTok{"blank"}\NormalTok{, }\StringTok{"congruent"}\NormalTok{,}\StringTok{"incongruent"}\NormalTok{)))}

\CommentTok{\# center TIME (period) on bin 9, and trial on trial 1000 and rescale; Add dummy variables for each bin.}
\NormalTok{ptb\_data }\OtherTok{\textless{}{-}}\NormalTok{ ptb\_data }\SpecialCharTok{\%\textgreater{}\%} 
        \FunctionTok{mutate}\NormalTok{(}\AttributeTok{period\_9 =}\NormalTok{ period }\SpecialCharTok{{-}} \DecValTok{9}\NormalTok{,}
               \AttributeTok{trial\_c =}\NormalTok{ (trial }\SpecialCharTok{{-}} \DecValTok{1000}\NormalTok{)}\SpecialCharTok{/}\DecValTok{1000}\NormalTok{,}
               \AttributeTok{d6  =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{6}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d7  =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{7}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d8  =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{8}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d9  =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{9}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d10 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{10}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d11 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{11}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d12 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{12}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d13 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{13}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d14 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{14}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{),}
               \AttributeTok{d15 =} \FunctionTok{if\_else}\NormalTok{(period }\SpecialCharTok{==} \DecValTok{15}\NormalTok{, }\DecValTok{1}\NormalTok{, }\DecValTok{0}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\subsection{Prior distributions}\label{prior-distributions}

To get the posterior distribution of the parameters given the data, we need to specify the prior distribution of the parameters. The middle column of Figure 4 shows seven examples of prior distributions on the logit and/or cloglog scales.

{\centering \includegraphics[width=0.8\linewidth,height=0.67\textheight,]{../Tutorial_2_Bayesian/figures/plot_of_priors}

\caption{Prior distributions on the logit and/or cloglog scales (middle column), and their implications on the probability scale after applying the inverse-logit (or logistic) transformation (left column), and the inverse-cloglog transformation (right column).}\label{fig:plot-priors}
\end{figure}

While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B in Figure 3 show that specifying such distributions on the logit and cloglog scales leads to rather informative distributions on the original probability (i.e., discrete-time hazard) scale, as most mass is pushed to probabilities of 0 and 1.

To gain a sense of what prior \emph{logit} values would approximate a uniform distribution on the probability scale, Kurz (2023a) simulated a large number of draws from the Uniform(0,1) distribution, converted those draws to the log-odds metric, and fitted a Student's t distribution. Row C in Figure 4 shows that using a t-distribution with 7.61 degrees of freedom and a scale parameter of 1.57 as a prior on the logit scale, approximates a uniform distribution on the probability scale. According to Kurz (2023a), such a prior might be a good prior for the intercept(s) in a logit-hazard model, while the N(0,1) prior in row D might be a good prior for the beta parameters in a logit-hazard model, as it gently regularizes p towards .5 (i.e., a zero effect on the logit scale).

To gain a sense of what prior \emph{cloglog} values would approximate a uniform distribution on the hazard probability scale, we followed Kurz's approach and simulated a large number of draws from the Uniform(0,1) distribution, converted them to the cloglog metric, and fitted a skew-normal model (due to the asymmetry of the cloglog link function). Row E shows that using a skew-normal distribution with a mean of -0.59, a standard deviation of 1.26, and a skewness of -4.22 as a prior on the cloglog scale, approximates a uniform distribution on the probability scale.
However, because hazard values below .5 are more likely in RT studies, using a skew-normal distribution with a mean of -1, a standard deviation of 1, and a skewness of -2 as a prior on the cloglog scale (row F), might be a good weakly informative prior for the alpha parameters or intercept(s) in a cloglog-hazard model.
A skew-normal distribution with a mean of -0.2, a standard deviation of 0.71, and a skewness of -2.2 might be a good weakly informative prior for the beta parameters in a cloglog-hazard model as it gently regularizes p towards .6321 (i.e., a zero effect on the cloglog scale).

@@

For the first model, we use a general specification of TIME (i.e., one intercept per time bin) for the baseline condition (blank prime), and assume that the effects of prime-target congruency and trial number are proportional and additive, and that the effect of trial number is linear.
Before we fit model 1, we remove unnecessary columns from the data, and specify our priors. In the code of Tutorial 2, this is accomplished as follows.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# remove unnecessary columns before fitting a model}
\NormalTok{M1\_data }\OtherTok{\textless{}{-}}\NormalTok{ ptb\_data }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{c}\NormalTok{(bl,tr,trial,period, period\_9,d9))  }

\CommentTok{\# Specify priors}
\NormalTok{priors\_M1 }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}0.2,0.71,{-}2.2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{),       }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d6"}\NormalTok{),}
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d7"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d8"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d10"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d11"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d12"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d13"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d14"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d15"}\NormalTok{),}
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"Intercept"}\NormalTok{),}
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"normal(0, 1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"sd"}\NormalTok{),        }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"lkj(2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"cor"}\NormalTok{)            }
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We can now estimate our first Bayesion regression model, as follows.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plan}\NormalTok{(multicore)}

\NormalTok{model\_M1 }\OtherTok{\textless{}{-}}
   \FunctionTok{brm}\NormalTok{(}\AttributeTok{data =}\NormalTok{ M1\_data,}
       \AttributeTok{family =} \FunctionTok{binomial}\NormalTok{(}\AttributeTok{link=}\StringTok{"cloglog"}\NormalTok{),}
\NormalTok{       event }\SpecialCharTok{|} \FunctionTok{trials}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{0} \SpecialCharTok{+}\NormalTok{ d6 }\SpecialCharTok{+}\NormalTok{ d7 }\SpecialCharTok{+}\NormalTok{ d8 }\SpecialCharTok{+}\NormalTok{ Intercept }\SpecialCharTok{+}\NormalTok{ d10 }\SpecialCharTok{+}\NormalTok{ d11 }\SpecialCharTok{+}\NormalTok{ d12 }\SpecialCharTok{+}\NormalTok{ d13 }\SpecialCharTok{+}\NormalTok{ d14 }\SpecialCharTok{+}\NormalTok{ d15 }\SpecialCharTok{+} 
\NormalTok{         condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{+}

\NormalTok{                   (d6 }\SpecialCharTok{+}\NormalTok{ d7 }\SpecialCharTok{+}\NormalTok{ d8 }\SpecialCharTok{+} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ d10 }\SpecialCharTok{+}\NormalTok{ d11 }\SpecialCharTok{+}\NormalTok{ d12 }\SpecialCharTok{+}\NormalTok{ d13 }\SpecialCharTok{+}\NormalTok{ d14 }\SpecialCharTok{+}\NormalTok{ d15 }\SpecialCharTok{+}\NormalTok{ condition }\SpecialCharTok{+}\NormalTok{ trial\_c }\SpecialCharTok{|}\NormalTok{ pid),}
       \AttributeTok{prior =}\NormalTok{ priors\_M1,}
       \AttributeTok{chains =} \DecValTok{4}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{, }\AttributeTok{iter =} \DecValTok{3000}\NormalTok{, }\AttributeTok{warmup =} \DecValTok{1000}\NormalTok{,}
       \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.999}\NormalTok{, }\AttributeTok{step\_size =} \FloatTok{0.04}\NormalTok{, }\AttributeTok{max\_treedepth =} \DecValTok{12}\NormalTok{),}
       \AttributeTok{seed =} \DecValTok{12}\NormalTok{, }\AttributeTok{init =} \StringTok{"0"}\NormalTok{,}
       \AttributeTok{file =} \StringTok{"Tutorial\_2\_Bayesian/models/model\_M1"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Estimating model\_M1 took about 70 minutes on a MacBook Pro (Sonoma 14.6.1 OS, 18GB Memory, M3 Pro Chip).

@@

For the second model, we use a third-order polynomial specification of TIME for the baseline condition (blank prime), and assume that the effects of prime-target congruency and trial number are proportional and additive, and that the effect of trial number is linear.

@@

\subsection{Advantages of hazard analysis}\label{advantages-of-hazard-analysis}

Statisticians and mathematical psychologists recommend focusing on the hazard function when analyzing time-to-event data for various reasons. First, as discussed by Holden, Van Orden, and Turvey (2009), ``probability density functions can appear nearly identical, both statistically and to the naked eye, and yet are clearly different on the basis of their hazard functions (but not vice versa). Hazard functions are thus more diagnostic than density functions'' (p.~331) when one is interested in studying the detailed shape of a RT distribution (see also Figure 1 in \ldots.).

@@

One important limitation is that you need a relatively large number of trials per condition to estimate the hazard function with high temporal resolution. There is a tradeoff between the number of trials per condition and the temporal resolution (i.e., bin width) of the hazard function. We therefore recommend to design as many trials as possible per experimental condition given the available resources.
Another limitation of multilevel hazard modeling is that it might take a long time to estimate the parameters.
